Comparison of several sequential and parallel derivative-free global optimization algorithms

### 2018-11-28 2019/PAVT/pavt_paper.tex
PAVT: initial english version

\documentclass{svproc}
%
% RECOMMENDED %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\usepackage{graphicx}
\usepackage{marvosym}
\usepackage{amssymb}
\usepackage{cite}
%%%%%ДОБАВИЛ ДЛЯ РУССКОГО ТЕКСТА
\usepackage[utf8x]{inputenc}
\usepackage[english,russian]{babel}
\usepackage{cmap}
%%%%%
% to typeset URLs, URIs, and DOIs
\usepackage{url}
\usepackage{hyperref}
\def\UrlFont{\rmfamily}

@@

\begin{abstract}
This work considers several stochastic and deterministic derivative-free global optimization algorithms. In the first part of the paper popular sequential open-source solvers are compared against the Globalizer solver, which is developed at the Lobachevsky State University. The Globalizer is designed to solve problems with black-box objectives satisfying the Lipschitz condition and shows competitive performance with other similar solvers. The comparison is done on several sets of challenging multi-extremal benchmark functions. The second part of this work is devoted to comparison between the Globalizer and MIDACO solvers on systems with shared and distributed memory. MIDACO is a state-of-the-art global solver included to the TOMLAB optimization environment for MATLAB. Results of the benchmark show advantages of the Globalizer on small-dimensional, but sufficiently multi-extremal benchmark functions.
% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\keywords{deterministic global optimization $\cdot$ stochastic global optimization
  $\cdot$ parallel numerical methods $\cdot$ derivative-free algorithms $\cdot$ black-box optimization}
\end{abstract}
%
\section{Introduction}

\begin{Russian}
Нелинейная глобальная оптимизация невыпуклых функций традиционно считается одной из самых трудных
задач математического программирования. Отыскание глобального минимума функции от нескольких переменных
зачастую оказывается сложнее, чем локальная оптимизация в тысячемерном пространстве. Для последней может оказаться достаточно
применения простейшего метода градиентного спуска, в то время как чтобы \textit{гаранитрованно} отыскать глобальный оптимум методам
оптимизации приходится накапливать информацию о поведении целевой функции во всей области поиска \cite{Jones2009,Paulavicius2011,Evtushenko2013,strSergGO}. В последнее время стали популярны
различные стохастические алгоритмы глобально оптимизации, прежде всего эволюционные \cite{Storn1997, SCHLUTER2009, KennedyEberhart1995}. Они имеют довольно простую структуру, позволяют решать задачи большой размерности, однако обеспечивают глобальную сходимость только в вероятностном смысле.

В данной работе рассмотрены open-source реализации девяти различных методов глобальной оптимизации.
Все алгоритмы были протестированы на наборе из 900 существенно многоэкстремальных функций, который был сгенерирован с
помощью специализированных генераторов задач \cite{Gaviano2003, grishaginClass}. Помимо сравнения последовательных алгоритмов, на подмножестве из 200 тестовых задач произведено сравнение солвера MIDACO \cite{Schlueter2012} и программной системы Globalizer \cite{globalizerSystem,Strongin2018} в условиях работы на суперкомпьютере Лобачевский.
\end{Russian}

@@

\begin{Russian}
Ранее в литературе рассмартивалось как сравнение между собой стохастических алгоритмов глобальной оптимизации \cite{Ali2005, JSSv060i06}, так и детерминированных \cite{posik2012, KVASOV2018245, Liberti2005}.
В них большинство современных методов изучены довольно детально, но упор в основном делается на последовательные алгоритмы, а основными критериями эффективности метода оптимизации является его надёжность и скорость сходимости. В большинстве работ в качестве набора тестовых функций берётся набор известных тестовых задач (например, функция Растригина, Ackley function и др.). Размер такого набора обычно не превышает 100 различных функций, некоторые из которых могут быть одноэкстремальны (как функция Розенброка).

В работе \cite{Beiranvand2017} сформулированы некие общие принципы, которых, по мнению авторов, следует придерживаться при сравнении методов оптимизации.
В частности, авторы говорят о преимуществах генераторов задач, позволяющих создать большие наборы задач, сводя случайные эффекты при сравнении к минимуму. В то же время, использование одного генератора может оказаться недостаточно для исчерпывающего сравнения методов. Чтобы частично преодолеть эту проблему, авторы \cite{Beiranvand2017} советуют использовать несколько генераторов различной природы и создавать наборы задач различной сложности.

Учитывая опыт предыдущих работ в области сравнения методов оптимизации, в данной работе будут использованы два генератора тестовых задач разной природы, с помощью которых сгенерировано 9 наборов по 100 задач разной сложности размерности от 2 до 5. Помимо сравнения последовательных методов, также в работе приводится сравнение эффективности двух параллельных алгоритмов.
\end{Russian}

@@

The convergence conditions and exact formulas for descision rules $R(i)$ and $d(x)$ of the
described algorithm are given, for example, in \cite{strSergGO}.

\subsection{Parallel Algorithm on Distributed Memory Exploiting a Set of Evolvents}
\label{sub:parallel_evolvents}
\subsubsection{Rotated Evolvents}
One of the possible ways to overcome the negative effects of using a numerical
approximation of evolvent (it destroys the information about the neighbor points in $\mathbb{R}^N$ space)
consists in using the multiple mappings
\begin{equation}
  Y_L(x)=\left\{y^0(x),\ y^1(x),...,\ y^L(x)\right\}
\end{equation}
instead of single Peano curve $y(x)$ (see \cite{strSergGO}).
The building of a set of Peano curves by rotation of the evolvents around the coordinate origin is a distinctive feature was proposed in \cite{Gergel2009}. Taking into account the initial mapping, one can conclude that current implementation of the
method allows to build up to $N(N-1)+1$ evolvents for mapping the $N$-dimensional domain
onto the corresponding one-dimensional intervals. This method for building a set of mappings can be ``scaled'' easily to obtain more evolvents (up to
$2^N$) if necessary.

Using the multiple mapping allows solving initial problem (\ref{eq:task}) by parallel solving the
problems
\[
\min\{\varphi(y^s(x)):x\in [0,1]\}, 1\leqslant s\leqslant S
\]
on a set of intervals $[0,1]$ by the index method. Each one-dimensional problem is solved on a
separate processor. The trial results at the point \(x^k\) obtained for the problem being solved by
particular processor are interpreted as the results of the trials in the rest problems (in the
corresponding points \(x^{k_1},\dots,x^{k_S})\). In this approach, a trial at the point \(x^k \in
[0,1]\) executed in the framework of the \(s\)-th problem, consists in the following sequence of
operations:
\par
1. Determine the image \(y^k=y^s (x^k)\) for the evolvent \(y^s (x)\).
\par
2. Inform the rest of processors about the start of the trial execution at the point \( y^k\) (the
blocking of the point \(y^k\) ).
\par
3. Determine the preimages \(x{}^{k_s}  \in [0,1], 1\leqslant s\leqslant S\), of the point \(y^k\) and interpret the
trial executed at the point \(y^k \in D \) as the execution of the trials in the \(S\) points
\(x{}^{k_1} ,\dots,x{}^{k_s} \)
\par
4. Inform the rest of processors about the trial results at the point \(y^k\).
\par
The decision rules for the mentioned parallel algorithm, in general, are the same as the rules of the
sequential algorithm (except the method of the trial execution). Each processor has its own copy
of the software realizing the computations of the problem functions and the decision rule of the
index algorithm. For the organization of the interactions among the processors, the queues are
created on each processor, where the processors store the information on the executed iterations
in the form of the tuples: the processor number \(s\), the trial point \(x{}^{k_s}\).
The modifications of the method taking into account multiply criteria are given in \cite{GERGEL2017,Gergel2018}.
\par
The mentioned parallelization scheme was proposed in \cite{Gergel2009} and implemented in the Globalizer system with the use of MPI technology. Main
features of implementation consist in the following:
\begin{itemize}
  \item A separate MPI-process is created for each
 of \(S\) one-dimensional problems being solved, usually, one process per one processor
 employed.
 \item Each process can use $p$ threads to parallel execute $p$ trials, usually one thread per an accessible core.
\end{itemize}

@@

  \item \textbf{Multi Level Single Linkage} \cite{Kan1987StochasticGO}. MLSL is an improved multistart algorithm.
  It samples low-discrepancy starting points and does local optimizations from them. In contrast to the dummy multistart schemes
  MLSL uses some clustering heuristics to avoid multiple local descents to already explored local minimas.

@@

  \item \textbf{Locally-based DIRECT (DIRECT$l$)} \cite{Gablonsky2001}. It's a variation of DIRECT which pays less attention to non-promising boxes and therefore
  has less exploration power: it can converge faster on problems with few local minimas, but lost the global one in complicated cases.

@@

  \item \textbf{StoGO} \cite{Madsen1998}. StoGO is dividing the search space into smaller hyper-rectangles via a branch-and-bound approach,
  and searching them by a local-search algorithm, optionally including some randomness.

@@

All the mentioned algorithms (except of AGS\(l\) which implemented in the Globalizer system)
are available in source codes as parts of wide-spread optimization packages.
AGS, DIRECT, DIRECT$l$, CRS, MLSL and StoGO are part of the NLOpt libray \cite{nlopt}. Differential Evolution and DSA can be found in
the latest version of the SciPy \cite{scipy} package for Python.

\section{Tools for Comparison of Global Optimization Algorithms}
\begin{Russian}

Использование сгенерированных некоторыми случайными механизмами
наборов тестовых задач с известными решениями является одним из общепринятых подходов
к сравнению алгоритмов оптимизации \cite{Beiranvand2017}. В данной работе
будем использовать два генератора тестовых задач, порождающих задачи различной природы \cite{grishaginClass, Gaviano2003}.

Обозначим набор задач, полученных с помощью первого генератора из \cite{grishaginClass} как \(F_{GR}\). Механизм построения задач \(F_{GR}\),
не предусматривает контроль за сложностью задач и количеством локальных оптимумов, однако известно, что порождаемые функции
являются существенно многоэкстремальными. Кроме того, задачи, порождаемые \(F_{GR}\) двухмерные. В данной работе будем использовать
100 случайно сгенерированных функций из класса \(F_{GR}\).

Генератор GKLS \cite{Gaviano2003} позволяет получать задачи заданной размерности и с заданным количеством экстремумов.
Кроме того GKLS позволяет регулировать сложность задач, уменьшая или увеличивая размер области притяжения глобального минимума.
В работе \cite{SergeyevKvasov2006} указаны параметры генератора, позволяющие породить наборы по 100 задач двух уровней сложности
(Simple и Hard) размерностей 2, 3, 4 и 5. Следуя примеру авторов генератора GKLS, будем использовать предложенные ими параметры и,
таким образом, добавим в тестовый ещё 800 задач различной размерности и сложности.

Будем считать, что тестовая задача решена, если метод оптимизации провёл очередное испытание \(y^k\) в
\(\delta\)-окрестности глобального минимума \(y^*\), т.е. $\left\|y^k-
y^*\right\|\leqslant \delta = 0.01\left\|b-a\right\|$, где \(a\) и \(b\) --- левая и правая границы гиперкуба из (\ref{eq:task}).
Если указанное соотношение не выполнено до истечения лимита на количество испытаний, то задача считается нерешённой.
Максимальный лимит на количество испытаний установлен для каждого класса задач, в соответствии с размерностью и сложностью (см. таблицу \ref{tab:limits}).

\begin{table}
\begin{center}
\caption{Trials limit for the each of test problems classes}
  \begin{tabular}{|l|{c}|}
    \hline
  Problems class & Trials limit\\
  \hline
  \(F_{GR}\) & 5000 \\
  \hline
  GKLS 2d Simple & 8000 \\
  \hline
  GKLS 2d Hard & 9000 \\
  \hline
  GKLS 3d Simple & 15000 \\
  \hline
  GKLS 3d Hard & 25000 \\
  \hline
  GKLS 4d Simple & 150000 \\
  \hline
  GKLS 4d Hard & 250000 \\
  \hline
  GKLS 5d Simple & 350000 \\
  \hline
  GKLS 5d Hard & 600000 \\
  \hline
  \end{tabular}
  \label{tab:limits}
\end{center}
\end{table}

В качестве характеристик метода оптимизации на каждом из классов будем рассматривать среднее число
испытаний, затраченное для решения одной задачи, и количество решённых задач. Чем меньше число испытаний, тем быстрее метод сходится
к решению, а значит и меньше обращается к потенциально трудоёмкой процедуре вычислений целевой функции.
Количество решённых задач говорит о надёжности метода с заданными параметрами на решаемом классе тестовых задач.
Чтобы сделать величины, характеризующие надёжность и скорость сходимости, независимыми при подсчёте среднего количества испытаний учитывались только решённые задачи.
\end{Russian}

\section{Results of Numerical Experiments}
\subsection{Results of Sequential Algorithms}
\begin{Russian}

Результаты тех или иных алгоритмов на различных классах задач напрямую зависят от
настроек алгоритмов. В большинстве случаев авторы программных реализаций ориентируются на
задачи средней сложности и чтобы получить удовлетворительный результат при решении
существенно многоэкстремальных задач, требуется корректировка некоторых параметров.
При проведении сравнения были заданы следующие параметры для методов:
\begin{itemize}
  \item в методе AGS\(l\) параметр чередования глобального и локального правил вычисления характеристик был задан равным 5:1;
  \item в методах DIRECT и DIRECT\(l\) параметр \(\epsilon=10^{-4}\);
  \item в методе SDA параметр \(visit=2.72\).
\end{itemize}

Остальные параметры варьировались в зависимости от класса задач (см. таблицу \ref{tab:params}).

@@

В таблицах \ref{tab:trials}, \ref{tab:solved} приведены результаты запуска методов оптимизации на рассматриваемых классах задач.
Методы DIRECT и AGS\(l\) показали наилучшую скорость сходимости на всех классах, причём AGS\(l\) уступает
DIRECT на задачах из классов Simple и обходит его на задачах классов Hard. Как видно из таблицы \ref{tab:solved},
самыми надёжными методами являются детерминированные: AGS, AGS\(l\), DIRECT и DIRECT\(l\). Из стохастических методов
самую высокую надёжность демонстрируют MLSL и SDA.

@@

The computational experiments with parallel algorithms have been carried out on the Lobachevsky supercomputer at
State University of Nizhni Novgorod. A computational node includes 2 Intel
Sandy Bridge E5-2660 2.2 GHz processors, 64 GB RAM. Each CPU has 8 cores (i. e. total 16
cores were available per a node). All computational nodes are working under CentOS Linux 7.2.
All considered algorithms are implemented using C++ and compiled by GCC 4.8.5.

Сравнение двух параллельных алгоритмов, реализованных в системах Globalizer и MIDACO
было проведено на четырёхмерных задачах GKLS, т.к. задачи большей размерности не поддерживаются
бесплатной версией MIDACO. Как и в случае с последовательными алгоритмами, для кадого класса задач
были изменены параметры методов. Для MIDACO параметр \(focus = -1\). В случае с Globalizer параметр
\(r\) был выбран следуя таблице \ref{tab:params}. С удвоением количества задействованных развёрток
этот параметр уменьшался на 0.3, поскольку увеличение количества развёрток увеличивает и надёжность метода из секции \ref{sub:parallel_evolvents}
(подробнее см. \cite{SOVRASOV2018}).

Если считать, что затраты на параллелизм пренебрежимо малы по сравнению с затратами на вычисление целевых функций в задачах оптимизации, то ускорение по времени от использования параллельного метода будет равно ускорению по итерациям. Однако, в действительности это предположение справедливо,
тлолько если вычисление значения целевой функции достаточно трудоёмко. Во всех численных экспериментах время вычисления целевой функции занимает примерно $2\times 10^{-3}$с.

In Table~\ref{tab:iterations}, an averaged number of iterations when solving 100 problems from
each considered class is presented.
The number of iterations is reduced considerably with increasing the number of nodes and the
number of threads on each node. В скобках также указано количество решённых задач в каждом классе.
Globalizer решает практически все задачи в обоих классах, в то время как MIDACO не решает многие задачи из класса GKLS 4d Hard.
Стоит ещё отметить, что с ростом числа узлов надёжность MIDACO падает на обоих классах задач, в то время как надёжность Globalizer остаётся стабильной.

@@

Среднее ускорение по итерациям и среднее ускорение по времени (в скобках) приведены в таблице \ref{tab:speedup}.
В первой строке таблицы приведено среднее количество итераций и среднее время решения задачи в
последовательном режиме (в скобках). MIDACO для при параллельных вычислениях использует
меньшее количество коммуникаций, поэтому показатели ускорения монотонно возрастают с ростом количества
вычислительных узлов. В Globalizer каждый узел должен переслать результаты проведения итерации
всем другим узлам и с увеличением количества узлов и количества рабочих потоков на каждом узле
эффективность распараллеливания падает. Это можно видеть в таблице \ref{tab:speedup}, когда при переходе от 4х
узлов к 8 ускорение по времени при использовании 16 потоков упало. В случае одного потока на узел
ускорение обоих методов оптимизации примерно одинаковое. Однако стоит заметить, что Globalizer выполняет
меньше испытаний и при \(p=1\) тратит гораздо меньше времени на решение всех задач.
С ростом количества узлов и потоков MIDACO догоняет Globalizer по среднему времени решения задачи,
однако надёжность при этом падает.

Аномальное сверхлинейное ускорение по времени
MIDACO на классе GKLS 2d Hard при использовании 8 и 12 узлов можно объяснить тем, что, согласно таблице
\ref{tab:iterations}, MIDACO решил больше задач в параллельном режиме, чем в последовательном, а значит, на
меньшем количестве задач работал до исчерпания лимита испытаний.

>>>>

\documentclass{svproc}
%
% RECOMMENDED
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%
\usepackage{graphicx}
\usepackage{marvosym}
\usepackage{amssymb}
\usepackage{cite}
% to typeset URLs, URIs, and DOIs
\usepackage{url}
\usepackage{hyperref}
\def\UrlFont{\rmfamily}

@@

\begin{abstract}
This work considers several stochastic and deterministic derivative-free global optimization
algorithms. In the first part of the paper popular sequential open-source solvers are compared against
the Globalizer solver, which is developed at the Lobachevskii State University. The Globalizer is
designed to solve problems with black-box objectives satisfying the Lipschitz condition and shows
competitive performance with other similar solvers. The comparison is done on several sets of
challenging multi-extremal benchmark functions. The second part of this work is devoted to
comparison between the Globalizer and MIDACO solvers on systems with shared and distributed
memory. MIDACO is a state-of-the-art global solver included to the TOMLAB optimization
environment for MATLAB. Results of the benchmark show advantages of the Globalizer on small-
dimensional, but sufficiently multi-extremal benchmark functions.
% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\keywords{deterministic global optimization $\cdot$ stochastic global optimization
  $\cdot$ parallel numerical methods $\cdot$ derivative-free algorithms $\cdot$ black-box
optimization}
\end{abstract}
%
\section{Introduction}

The nonlinear global optimization of the non-convex functions is considered to be one of the most
difficult problems of mathematical programming traditionally. The finding of the global minimum of
a function of several variables appears to be more difficult than the local optimization in the
thousand-dimensional space often. For the latter, the application of the simplest gradient descent
method may appear to be enough whereas in order to \textit{guarantee} the finding of the global
optimum by the optimization methods, one has to accumulate the information on the behavior of the
objective function within the whole search domain
\cite{Jones2009,Paulavicius2011,Evtushenko2013,strSergGO}. To date, various stochastic global
optimization algorithms have attracted much attention, first of all, the evolution ones
\cite{Storn1997, SCHLUTER2009, KennedyEberhart1995}. These ones have rather simple structure
and allow solving the problems of large dimensionality. However, the evolution algorithms ensure
the global convergence in the probabilistic sense only.

In the present paper, the open source implementations of nine different global optimization methods
were considered. All algorithms were tested using a set of 900 essentially multiextremal functions,
which has been generated using special problem generators \cite{Gaviano2003, grishaginClass}.
Besides the comparison of the sequential algorithms, a comparison of the MIDACO solver
\cite{Schlueter2012} with the Globalizer software system \cite{globalizerSystem,Strongin2018} has been
performed while running on Lobachevsky supercomputer using a subset of 200 test problems.

@@

Earlier, the comparison of the stochastic global optimization algorithms \cite{Ali2005, JSSv060i06}
as well as of the deterministic ones \cite{posik2012, KVASOV2018245, Liberti2005} between each
other has been considered in the literature. In these works, the majority of modern methods have
been studied in details, but the emphasis was made mainly on the sequential algorithms while the
reliability and the convergence speed were the main criteria of efficiency of the optimization
methods. In the majority of works, the sets of well-known test problems (for example, the Rastrigin
functions, Ackley function, etc.) were taken as the sets of test functions. The sizes of such sets don't
exceed 100 different functions usually, some of which can be the single-extremal ones (such as the
Rosenbrock functions).

In Ref. \cite{Beiranvand2017}, some general principles were formulated, which, in the author's
opinion, should be obeyed when comparing the optimization methods. In particular, the authors say
about the advantages of the problem generators allowing generating the large sets of problems thus
minimizing the random effects when comparing the methods. At the same time, the use of a single
generator can appear to be not enough for a comprehensive comparison of the methods. In order to
overcome this problem in part, the authors of Ref. \cite{Beiranvand2017} advise to use several
generators of various nature and to create the sets of problems of various complexity.

Taking into account the experience of the preceding works in the field of comparison of the
optimization methods, two generators of the test problems of different nature will be used in the
present work. Using these ones, 9 sets of 100 problems of various complexity with the
dimensionality varying from 2 to 5 were generated. Besides the comparison of the sequential
methods, a comparison of the efficiencies of two parallel algorithms is presented in the work as well.

@@

The convergence conditions and exact formulas for decision rules $R(i)$ and $d(x)$ of the
described algorithm are given, for example, in \cite{strSergGO}.

\subsection{Parallel Algorithm on Distributed Memory Exploiting a Set of Evolvents}
\label{sub:parallel_evolvents}
\subsubsection{Rotated Evolvents}
One of the possible ways to overcome the negative effects of using a numerical
approximation of evolvent (it destroys the information about the neighbor points in
$\mathbb{R}^N$ space)
consists in using the multiple mappings
\begin{equation}
  Y_L(x)=\left\{y^0(x),\ y^1(x),...,\ y^L(x)\right\}
\end{equation}
instead of single Peano curve $y(x)$ (see \cite{strSergGO}).
The building of a set of Peano curves by rotation of the evolvents around the coordinate origin is a
distinctive feature was proposed in \cite{Gergel2009}. Taking into account the initial mapping, one
can conclude that current implementation of the
method allows to build up to $N(N-1)+1$ evolvents for mapping the $N$-dimensional domain
onto the corresponding one-dimensional intervals. This method for building a set of mappings can be
''scaled'' easily to obtain more evolvents (up to
$2^N$) if necessary.

Using the multiple mapping allows solving initial problem (\ref{eq:task}) by parallel solving the
problems
\[
\min\{\varphi(y^s(x)):x\in [0,1]\}, 1\leqslant s\leqslant S
\]
on a set of intervals $[0,1]$ by the index method. Each one-dimensional problem is solved on a
separate processor. The trial results at the point \(x^k\) obtained for the problem being solved by
particular processor are interpreted as the results of the trials in the rest problems (in the
corresponding points \(x^{k_1},\dots,x^{k_S})\). In this approach, a trial at the point \(x^k \in
[0,1]\) executed in the framework of the \(s^{th}\) problem, consists in the following sequence
of
operations:
\par
1. Determine the image \(y^k=y^s (x^k)\) for the evolvent \(y^s (x)\).
\par
2. Inform the rest of processors about the start of the trial execution at the point \( y^k\) (the
blocking of the point \(y^k\) ).
\par
3. Determine the preimages \(x{}^{k_s}  \in [0,1], 1\leqslant s\leqslant S\), of the point \(y^k\) and
interpret the
trial executed at the point \(y^k \in D \) as the execution of the trials in the \(S\) points
\(x{}^{k_1} ,\dots,x{}^{k_s} \)
\par
4. Inform the rest of processors about the trial results at the point \(y^k\).
\par
The decision rules for the mentioned parallel algorithm, in general, are the same as the rules of the
sequential algorithm (except the method of the trial execution). Each processor has its own copy
of the software realizing the computations of the problem functions and the decision rule of the
index algorithm. For the organization of the interactions among the processors, the queues are
created on each processor, where the processors store the information on the executed iterations
in the form of the tuples: the processor number \(s\), the trial point \(x{}^{k_s}\).
The modifications of the method taking into account multiply criteria are given in \cite{GERGEL2017,Gergel2018}.
\par
The mentioned parallelization scheme was proposed in \cite{Gergel2009} and implemented in the
Globalizer system with the use of MPI technology. Main
features of implementation consist in the following:
\begin{itemize}
  \item A separate MPI-process is created for each
 of \(S\) one-dimensional problems being solved, usually, one process per one processor
 employed.
 \item Each process can use $p$ threads to parallel execute $p$ trials, usually one thread per an
accessible core.
\end{itemize}

@@

  \item \textbf{Multi Level Single Linkage} \cite{Kan1987StochasticGO}. MLSL is an improved
multistart algorithm.
  It samples low-discrepancy starting points and does local optimizations from them. In contrast to
the dummy multistart schemes
  MLSL uses some clustering heuristics to avoid multiple local descents to already explored local
minima.

@@

  \item \textbf{Locally-based DIRECT (DIRECT$l$)} \cite{Gablonsky2001}. It's a variation of
DIRECT which pays less attention to non-promising boxes and therefore
  has less exploration power: it can converge faster on problems with few local minima, but lost the
global one in complicated cases.

@@

  \item \textbf{StoGO} \cite{Madsen1998}. StoGO is dividing the search space into smaller hyper-
rectangles via a branch-and-bound approach,
  and searching them by a local-search algorithm, optionally including some randomness.

@@

All the mentioned algorithms (except of AGS\(l\) which implemented in the Globalizer system)
are available in source codes as parts of wide-spread optimization packages.
AGS, DIRECT, DIRECT$l$, CRS, MLSL and StoGO are part of the NLOpt library \cite{nlopt}.
Differential Evolution and DSA can be found in
the latest version of the SciPy \cite{scipy} package for Python.

\section{Tools for Comparison of Global Optimization Algorithms}

The use of the sets of test problems with known solutions generated by some random mechanisms is
one of commonly accepted approaches to comparing the optimization algorithms
\cite{Beiranvand2017}. In the present work, we will use two generators of test problems generating
the problems of different nature \cite{grishaginClass, Gaviano2003}.

Let us denote the problem set obtained with the use of the first generator from \cite{grishaginClass}
as \(F_{GR}\). The mechanism of generation of the problems \(F_{GR}\) doesn't provide the
control of the problem complexity and of the number of local optima. However, the generated
functions are known to be the multiextremal ones essentially. Besides, the problems generated by
\(F_{GR}\) are the two-dimensional ones. In the present work, we will use 100 functions from the
class \(F_{GR}\) generated randomly.

The GKLS generator \cite{Gaviano2003} allows obtaining the problems of given dimensionality
with given number of extrema. Besides, GKLS allows adjusting the complexity of the problems by
decreasing or increasing the size of the global minimum attractor. In Ref.
\cite{SergeyevKvasov2006}, the parameters of the generator allowing generating the sets of 100
problems each of two levels of complexity (Simple and Hard) of the dimensionality equal to 2, 3, 4,
and 5 are given. Following the authors of the GKLS generator, we will use the parameters proposed
by them and, this way, add 800 more problems of various dimensionalities and complexity into the
test problem set.

Let us suppose a test problem to be solved if the optimization method executes the scheduled trial
\(y^k\) in a \(\delta\)-nearness of the global minimum \(y^*\) i. e. $\left\|y^k-y^*\right\|\leqslant \delta
= 0.01\left\|b-a\right\|$ where \(a\) and \(b\) are the left and the right boundaries of the hypercube
from (\ref{eq:task}). If this relation is not fulfilled before the expiration of the limit of the number of
trials, the problem was considered to be unsolved. The maximum limit of the number of trials was set
for each problem class according to the dimensionality and complexity (see Table \ref{tab:limits}).

\begin{table}
\begin{center}
\caption{Trials limits for the test problem classes}
  \begin{tabular}{|l|{c}|}
    \hline
  Problems class & Trials limit\\
  \hline
  \(F_{GR}\) & 5000 \\
  \hline
  GKLS 2d Simple & 8000 \\
  \hline
  GKLS 2d Hard & 9000 \\
  \hline
  GKLS 3d Simple & 15000 \\
  \hline
  GKLS 3d Hard & 25000 \\
  \hline
  GKLS 4d Simple & 150000 \\
  \hline
  GKLS 4d Hard & 250000 \\
  \hline
  GKLS 5d Simple & 350000 \\
  \hline
  GKLS 5d Hard & 600000 \\
  \hline
  \end{tabular}
  \label{tab:limits}
\end{center}
\end{table}

Let us consider the averaged number of trials executed to solve a single problem and the number of
solved problems as the characteristics of the optimization method on each class. The less the number
of trials, the faster the method converges to a solution, hence the less times it turns to a potentially
computation-costly procedure of computing the objective function. The number of solved problems
evidences the reliability of the method at given parameters on the class of test problems being
solved. In order to calculate the quantities featuring the reliability and the speed of convergence,
when calculating the averaged number of trials the solved problems were taken into account only.

\section{Results of Numerical Experiments}
\subsection{Results of Sequential Algorithms}

The results of various algorithms on different problem classes depend on the adjustments of
algorithms directly. In most cases, the authors of software implementations are oriented onto the
problems of medium difficulty. In order to obtain a satisfactory result when solving the essentially
multiextremal problems, a correction of some parameters is required. When conducting the
comparison, the following parameters for the methods were employed:
\begin{itemize}
  \item in the AGS\(l\) method, the parameter of alternation of the global rules of computing the
characteristic and the local one was set to be equal to 5:1;
  \item in the DIRECT and DIRECT\(l\) methods, the parameter \(\epsilon=10^{-4}\);
  \item in the SDA method, the parameter \(visit=2.72\).
\end{itemize}

The rest parameters were varied subject to the problem class (see Table \ref{tab:params}).

@@

The results of running the optimization methods on the considered problem classes are presented in
Tables \ref{tab:trials}, \ref{tab:solved}. The DIRECT and AGS\(l\) methods have demonstrated the
best convergence speed on all classes, at that AGS\(l\) inferior to DIRECT on the problems from the
Simple classes and has an advantage on the problems of the Hard classes. As one can see from Table
\ref{tab:solved}, the deterministic methods (AGS, AGS\(l\), DIRECT, and DIRECT\(l\)) were the
most reliable. Among the stochastic methods, MLSL and SDA have demonstrated the highest
reliability.

@@

The computational experiments with parallel algorithms have been carried out on the Lobachevskii
supercomputer at
State University of Nizhni Novgorod. A computational node includes 2 Intel
Sandy Bridge E5-2660 2.2 GHz processors, 64 GB RAM. Each CPU has 8 cores (i. e. total 16
cores were available per a node). All computational nodes are working under CentOS Linux 7.2.
All considered algorithms are implemented using C++ and compiled by GCC 4.8.5.

The comparison of two parallel algorithms implemented in the Globalizer and MIDACO systems
was conducted on four-dimensional GKLS problems since the problems of larger dimensionality are
not supported by the free version of MIDACO. As in the case of the sequential algorithms, the
parameters of methods were determined for each problem class. For MIDACO, the parameter
\(focus = -1\). In the case of Globalizer, the parameter \(r\) was selected following Table
\ref{tab:params}. This parameter decreased by 0.3 with doubling the number of employed evolvents
since the increasing of the number of evolvents enhances the reliability of the method from Section
\ref{sub:parallel_evolvents} as well (see \cite{SOVRASOV2018} for details).

If one considers the costs of parallelism to be negligible as compared to the costs of computing the
objective functions in the optimization problems, the speedup in time due to the use of the parallel
method would be equal to the speedup with respect to the number of iterations. However, actually
this assumption is true if the computing of the objective function is well computation costly only. In
all numerical experiments, the time of computing the objective function was approximately $2.5\times
10^{-3}$ s.

In Table~\ref{tab:iterations}, an averaged number of iterations when solving 100 problems from
each considered class is presented.
The number of iterations is reduced considerably with increasing the number of nodes and the
number of threads on each node. Also, the number of solved problems in each class is given  in
brackets.
Globalizer solves almost all problems in both classes whereas MIDACO hasn't solved a considerable
number of problems from the GKLS 4d Hard class. Also, it is worth noting that the reliability of
MIDACO decreased with increasing number of nodes on both problem classes whereas the
reliability of Globalizer remains stable.

@@

The averaged speedup in the number of iterations and the one in time (in brackets) are presented in
Table \ref{tab:speedup}. In the first row of Table \ref{tab:speedup}, the averaged number of
iterations and the averaged time of solving the problems in the sequential mode (in brackets) are
presented. MIDACO uses a smaller number of communications in the parallel computations.
Therefore, the indicators of speedup increased monotonously with increasing number of
computational nodes employed. In Globalizer, each node should transfer the results of the executed
iterations to all other nodes. Therefore, the efficiency of parallelization decreased with increasing
number of nodes and the number of active threads on each node. This is evident from Table
\ref{tab:speedup}: the speedup in time decreased at the transition from 4 nodes to 8 ones when
using 16 threads. In the case of using 1 thread per a node, the speedup of both optimization methods
were approximately the same. However, it should be noted that Globalizer executed less number of
trials, and spends much less time for solving all problems when \(p=1\). With increasing number of
nodes and threads, MIDACO approaches Globalizer in the averaged time of problem solving,
however, its reliability decreases at the same time.

An anomalous superlinear speedup of MIDACO in time on the GKLS 2d Hard class when
employing 8 and 12 nodes can be explained by the fact that, according to Table \ref{tab:iterations},
MIDACO has solved a greater number of problems in the parallel mode than in the sequential one,
hence, has worked until the limit of trials is expired on a less number of problems.
